---
title: "STAT204 - HW01"
format: pdf
---
# Setup
## Loading Relevant Package

```{r}
library(palmerpenguins)
```

## Exploring "penguin" Data File

```{r}
summary(penguins)

## to check how many rows are in the dataframe
nrow(penguins)

## to check how many columns are in the dataframe
ncol(penguins)

## to check what variable types are in the dataframe
str(penguins)
```

# Part 1: Vectors and Summary Statisitcs
## 1.1 Extract and Analyze Bill Lengths

```{r}
## Extract bill_length_mm as a vector
## this pulls out the data in the "bill_length_mm" column
bill_lengths <- penguins$bill_length_mm

## Calculate: mean, median, sd, min, max (handle NAs!)
mean_bill_length <- mean(bill_lengths, na.rm = TRUE)
median_bill_length <- median(bill_lengths, na.rm = TRUE)
sd_bill_length <- sd(bill_lengths, na.rm = TRUE)
min_bill_length <- min(bill_lengths, na.rm = TRUE)
max_bill_length <- max(bill_lengths, na.rm = TRUE)
```

## 1.2 Logical Subsetting

```{r}
# Create logical vector: TRUE when bill length > 45mm
long_bills <- penguins$bill_length_mm > 45

# How many penguins have long bills?
sum(long_bills, na.rm = TRUE) ## 165 penguins have bills longer that 45mm
```

# Part 2: Subsetting Data 
## 2.1 Island analysis 

```{r}
## to check how many differnt island names are in the column
unique(penguins$island)

# Extract bill lengths for Dream island penguins only
dream_bill_lengths <- penguins$bill_length_mm[penguins$island == "Dream"]
dream_bill_lengths

# Calculate the mean for this subset
mean(dream_bill_lengths, na.rm = TRUE)
```

## 2.2 Species Counting

```{r}
# How many Adelie penguins are in the dataset?
adelie_count <- sum(penguins$species == "Adelie")
adelie_count
```

## 2.3 Find the Heaviest Penguin

```{r}
# Find: body mass, species, and island of the heaviest penguin

## to identify row index (or location) of the maximum body mass in the dataframe
big_penguin_row <- which.max(penguins$body_mass_g) 
big_penguin_row

## extract information from specific columns in the row
penguins[big_penguin_row, c("body_mass_g", "species", "island")]
```

# Part 3: Matrices and apply()
## 3.1 Create Measurement Matrix

```{r}
# Remove rows with missing values
penguins_complete <- penguins[complete.cases(penguins), ]

# Create matrix with: bill_length_mm, bill_depth_mm, 
# flipper_length_mm, body_mass_g

## column bind function to combine vectors side by side as columns in a matrix   
penguin_measurements <- cbind(
    penguins$bill_length_mm,
    penguins$bill_depth_mm, 
    penguins$flipper_length_mm,
    penguins$body_mass_g
)

## column names assigns names to columns in new matrix
colnames(penguin_measurements) <- c(
    "bill_length_mm",
    "bill_depth_mm",
    "flipper_length_mm",
    "body_mass_g"
)

```

## 3.2 Apply Functions

```{r}
# Use apply() to calculate mean of each column
apply(penguin_measurements, 2, mean, na.rm = TRUE)  ## the number 2 is for columns

# Use apply() to find which row has the maximum for each column
## which.max finds position of max value
## note this is not the actual max value in each column
apply(penguin_measurements, 2, which.max)  
```

# Part 4: Custom Function

```{r}
# Function that: removes NAs, returns (x - mean) / sd
## this creates a standardizing function called "standardize"

standardize <- function(x) {
  x <- na.omit(x)             # this removes NAs
  (x - mean(x)) / sd(x)       # this standardizes each value
}
 
# Test on bill_length_mm
bill_length_standardized <- standardize(penguins$bill_length_mm)

# Verify: mean ≈ 0, sd ≈ 1
mean(bill_length_standardized)
sd(bill_length_standardized)
```

## Bonus
**Option A**: Create a summary data frame with one row per species showing: count, mean body mass, mean flipper length

```{r}
## first need to create 4 vectors

## 1. to get the unique number of species 
species <- unique(penguins$species)

## 2. to count the number of penguins in each species
penguin_count <- table(penguins$species)

## 3. to calculate mean body mass for each species
mean_body_mass <- tapply(
    penguins$body_mass_g,   ## to specify the variable
    penguins$species,       ## to specify how it should be grouped (e.g. by species)
    mean,                   ## to specify desired function
    na.rm = TRUE            ## to remove NAs
)

## 4. to calculate mean flipper length for each species 
mean_flipper_length <- tapply(
    penguins$flipper_length_mm, ## to specify the variable
    penguins$species,           ## to specify how it should be grouped 
    mean,                       ## to specify desired function
    na.rm = TRUE                ## to remove NAs
)

## summary dataframe
penguins_summary_df <- data.frame(
    species = species,                ## left side is the name of column
    count = penguin_count,            ## right side is the object containing the data
    mean_body_mass = mean_body_mass,  
    mean_flipper_length = mean_flipper_length
)
```